---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- To generate README.md for GitHub homepage: copy this file to the package root directory, set for_github parameter to TRUE, then knit. Delete the README.Rmd file from the root directory after completion -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  # specify the directory where figures generated by code chunks in an R Markdown document should be saved
  fig.path = "man/figures/README-", 
  out.width = "100%"
)
```


# Consistent annotations for all fly motif database from MEME Suite

## Problem

[MEME Suite](https://meme-suite.org/meme/doc/download.html) provided 5 motif database for drosophila, but each database has it's unique motif id naming, which makes it hard to related it to the gene name and make a comparision between database.


```
$ tree
|-- OnTheFly_2014_Drosophila.meme
|-- dmmpmm2009.meme
|-- fly_factor_survey.meme
|-- flyreg.v2.meme
`-- idmmpmm2009.meme
```


```
$ ls *meme | while read id; do echo $id; grep 'MOTIF' $id | head -n 5; done

OnTheFly_2014_Drosophila.meme
MOTIF OTF0001.1 7UP1_DROME_B1H
MOTIF OTF0002.1 A0AQF9_DROME_B1H
MOTIF OTF0003.1 A0JQ60_DROME_SELEX
MOTIF OTF0003.2 A0JQ60_DROME_DNaseI
MOTIF OTF0004.1 A1A6R5_DROME_B1H

dmmpmm2009.meme
MOTIF abd-A
MOTIF Antp
MOTIF ap
MOTIF bcd
MOTIF brk

fly_factor_survey.meme
MOTIF FBgn0000014 AbdA_Cell
MOTIF FBgn0000014_2 Abd-A_FlyReg
MOTIF FBgn0000014_3 AbdA_SOLEXA
MOTIF FBgn0000015 AbdB_Cell
MOTIF FBgn0000015_2 Abd-B_FlyReg

flyreg.v2.meme
MOTIF abd-A
MOTIF Abd-B
MOTIF Adf1
MOTIF Aef1
MOTIF Antp

idmmpmm2009.meme
MOTIF abd-A
MOTIF Abd-B
MOTIF Antp
MOTIF bab1
MOTIF bcd
```

## Results

all ids from motif database has been related to it's latest FBgn and Symbol in Flybase. see ``Consistent-annotation-for-all-fly-motif-database.csv``

- source： database

- raw： raw motif entry

- FBgn: latest FBgn ID

- Symbol: latest gene symbol

And total 334 TFs with motifs are stored in all database. Is all these 334 genes are TFs?

```{r}
df <- read.csv('./Consistent-annotation-for-all-fly-motif-database.csv',stringsAsFactors = FALSE,row.names = 1)
for (i in unique(df$source)) {
    print(head(df[df$source == i,]))
}
```


```{r}
table(df$source)
length(unique(df$FBgn))
sort(table(df$Symbol),decreasing = TRUE)
```


## codes and files used

**files:**

``fb_synonym_fb_2026_01.tsv.gz`` downloaded from flybase which stores the latest FBgn and Symbol.

**codes:**

- 0_extract-motif-names.sh

- 1_Standardize-gene-names.R: update genes info for each database

- 2_merge-all.R: merge all results

## Thanks

Thanks to Claude code for generating the web crawler scripts.

