---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->
<!-- To generate README.md for GitHub homepage: copy this file to the package root directory, set for_github parameter to TRUE, then knit. Delete the README.Rmd file from the root directory after completion. -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  # specify the directory where figures generated by code chunks in an R Markdown document should be saved
  fig.path = "man/figures/README-", 
  out.width = "100%"
)
```


# Consistent annotations for all fly motif databases from MEME Suite

## Problem

The [MEME Suite](https://meme-suite.org/meme/doc/download.html) provides 5 motif databases for Drosophila, but each database has its own unique motif ID naming convention. This makes it difficult to relate motifs to gene names and to make comparisons between databases.


```
$ tree
|-- OnTheFly_2014_Drosophila.meme
|-- dmmpmm2009.meme
|-- fly_factor_survey.meme
|-- flyreg.v2.meme
`-- idmmpmm2009.meme
```


```
$ ls *meme | while read id; do echo $id; grep 'MOTIF' $id | head -n 5; done

OnTheFly_2014_Drosophila.meme
MOTIF OTF0001.1 7UP1_DROME_B1H
MOTIF OTF0002.1 A0AQF9_DROME_B1H
MOTIF OTF0003.1 A0JQ60_DROME_SELEX
MOTIF OTF0003.2 A0JQ60_DROME_DNaseI
MOTIF OTF0004.1 A1A6R5_DROME_B1H

dmmpmm2009.meme
MOTIF abd-A
MOTIF Antp
MOTIF ap
MOTIF bcd
MOTIF brk

fly_factor_survey.meme
MOTIF FBgn0000014 AbdA_Cell
MOTIF FBgn0000014_2 Abd-A_FlyReg
MOTIF FBgn0000014_3 AbdA_SOLEXA
MOTIF FBgn0000015 AbdB_Cell
MOTIF FBgn0000015_2 Abd-B_FlyReg

flyreg.v2.meme
MOTIF abd-A
MOTIF Abd-B
MOTIF Adf1
MOTIF Aef1
MOTIF Antp

idmmpmm2009.meme
MOTIF abd-A
MOTIF Abd-B
MOTIF Antp
MOTIF bab1
MOTIF bcd
```

## Why This Matters

Cross-database motif analysis is essential for understanding transcriptional regulation in *Drosophila*, but has been hindered by inconsistent naming conventions. This project addresses three critical challenges:

**1. Data Integration**
Different databases use completely different identification systems:
- `OnTheFly` uses internal OTF IDs with UniProt names and experimental methods (e.g., `OTF0001.1 7UP1_DROME_B1H`)
- `dmmpmm2009` and `flyreg.v2` use gene symbols directly (e.g., `abd-A`, `br-Z1`)
- `fly_factor_survey` combines FlyBase IDs with experimental metadata (e.g., `FBgn0000014_2 Abd-A_FlyReg`)

**2. Annotation Consistency**
The pipeline standardizes all motif entries using:
- Official FlyBase synonym mappings (`fb_synonym_fb_2026_01.tsv.gz`)
- Automated resolution of isoform variants (e.g., `br-Z1/Z2/Z3/Z4` → `br`)
- Updates to deprecated FBgn IDs via FlyBase conversion API
- Manual curation for ambiguous cases

**3. Cross-Database Analysis**
The unified annotation enables researchers to:
- Compare motif characteristics across multiple experimental datasets
- Identify overlapping and unique transcription factors across databases
- Conduct meta-analyses combining data from different sources
- Build comprehensive transcription factor regulatory networks

## Results

### Motif IDs maping file

All motif IDs from the databases have been linked to their latest FBgn and Symbol in FlyBase. See `Consistent-annotation-for-all-fly-motif-database.csv`.

- `source`: database name

- `raw`: raw motif entry

- `FBgn`: latest FBgn ID

- `Symbol`: latest gene symbol

A total of 334 TFs with motifs are stored across all databases. Are all these 334 genes truly TFs?

```{r}
df <- read.csv('./Consistent-annotation-for-all-fly-motif-database.csv',stringsAsFactors = FALSE,row.names = 1)
for (i in unique(df$source)) {
    print(head(df[df$source == i,]))
}
```


```{r}
table(df$source)
length(unique(df$FBgn))
sort(table(df$Symbol),decreasing = TRUE)
```

### Updated meme files with new MOTIF name

new MOTIF name: ``database name`` + ``FBgn ID`` + ``Gene Symbol`` + ``Row number``, separated by underscore.

Row number: the row number in ``Consistent-annotation-for-all-fly-motif-database.csv`` file, included to avoid duplicated motif names.

```
$ grep 'MOTIF' dmmpmm2009.meme |head -n 10
MOTIF idmmpmm2009.meme_FBgn0000014_abd-A_773
MOTIF idmmpmm2009.meme_FBgn0260642_Antp_775
MOTIF flyreg.v2.meme_FBgn0267978_ap_703
MOTIF idmmpmm2009.meme_FBgn0000166_bcd_777
MOTIF idmmpmm2009.meme_FBgn0024250_brk_778
MOTIF flyreg.v2.meme_FBgn0283451_br_760
MOTIF flyreg.v2.meme_FBgn0283451_br_761
MOTIF flyreg.v2.meme_FBgn0283451_br_762
MOTIF flyreg.v2.meme_FBgn0283451_br_763
MOTIF flyreg.v2.meme_FBgn0011723_byn_710
```

## Code and Files Used

**Files:**

`fb_synonym_fb_2026_01.tsv.gz` — downloaded from FlyBase, containing the latest FBgn and Symbol mappings.

**Code:**

- `0_extract-motif-names.sh` — extracts motif names from databases

- `1_Standardize-gene-names.R` — updates gene information for each database

- `2_merge-all.R` — merges all results

- `3_update-motif-names-in-meme-files.py` - update motif names in meme files

## Acknowledgments

Thanks to Claude Code for generating the web crawler scripts.

